---
title: "cdev value as a function of data characteristics"
output: html_notebook
---

How does the value of cdev depend on different characteristics of the data set, such as the level of inter-group variation, the number of genes, the percentage of DE genes? We'll use simulation to explore those relationships.


```{r setup, include = FALSE}
library(magrittr)
library(ggplot2)
library(parallel)
devtools::load_all('../')

logFC = function(x, groups, log.base = 2, pseudocount = 0) {
    G = unique(groups)
    log(apply(x[,groups == G[2]], MARGIN = 1, mean) + pseudocount, base = log.base) - 
        log(apply(x[,groups == G[1]], MARGIN = 1, mean) + pseudocount, base = log.base)
}
THEME_BASE_SIZE = 13
```

## Varying number of genes

```{r}
NSAMPLES = 40
nGenes = c(2000,10000, 20000, 50000)
percentDE = 0.5 
foldChange = 4
sims_g <- lapply(nGenes, function(nGene) {
    simulate.counts(nGene, NSAMPLES/2,NSAMPLES/2,fc = foldChange, percentDE = percentDE, completelyAsymm = TRUE)
}) %>%
    set_names(as.character(nGenes))

sims_g.ref = lapply(nGenes, function(nGene) {
    sim = sims_g[[as.character(nGene)]]
    normalize.by.refs(sim$counts, which(sim$differential == 0))
}) %>%
    set_names(as.character(nGenes))
```

```{r}
n_data_points = 30

sims_g.rand.norms =
    lapply(nGenes, function(nGene) {
        sim = sims_g[[as.character(nGene)]]
        sim.norm.ref = sims_g.ref[[as.character(nGene)]]
        trueRef.idx = which(sim$differential == 0)
        trueDif.idx = which(sim$differential == 1)
        NSCALE = min(length(trueDif.idx), length(trueRef.idx))
        mclapply(seq(1,NSCALE, floor(NSCALE/n_data_points)), mc.cores = 1, FUN = function(n_diff) {
            lapply(1:10, function(i) {
                rand.dif = sample(trueDif.idx, size = n_diff)
                rand.ref = sample(trueRef.idx, size = NSCALE - n_diff)
                rand.subset = union(rand.ref, rand.dif)
                raw = colSums(sim$counts[rand.subset,,drop= FALSE])
                sim.norm.i = normalize.by.refs(sim$counts, rand.subset)
                scalingFactors = raw / geom.mean(raw)
                
                return(list('scalingFactors' = scalingFactors,
                            'nGene' = nGene,
                            'rand.subset' = rand.subset,
                            'diffFraction' = n_diff/NSCALE,
                            'n_diffs_in_scale' = n_diff,
                            'diffVarTotal' = sum(var(sim$counts[rand.dif,])), # The total variance of the DE genes in the random set selected for scaling
                            'cdev' =  cdev(sim.norm.i, sim.norm.ref)))
            }) 
        }) %>%
        unlist(recursive = FALSE)
    }) %>%
    unlist(recursive = FALSE)
```

### DE tests

```{r, message=FALSE}
g_de.results = lapply(sims_g.rand.norms, function(sim.norm) {
    
    sim = sims_g[[as.character(sim.norm$nGene)]]
    de.test.limma(sim$counts,
            group = sim$conditions$condition,
            norm.factors = sim.norm$scalingFactors) %>%
        limma::topTable(n = Inf, sort.by = 'none', adjust.method = 'BH')
})
```

```{r,fig.width=15,fig.height=4}
performance.df = lapply(1:length(sims_g.rand.norms), function(i) {
    
    sim.norm = sims_g.rand.norms[[i]]
    sim = sims_g[[as.character(sim.norm$nGene)]]
    de = g_de.results[[i]]
    deg.idx = which(de$adj.P.Val < 0.01)
    auc = auroc(1-de$adj.P.Val, !!sim$differential)
    TP = length(intersect(which(sim$differential == 1), deg.idx))
    FP = length(setdiff(deg.idx, which(sim$differential == 1)))
    idx.diff = sim.norm$rand.subset[which(sim$differential[sim.norm$rand.subset] == 1)]
    
    data.frame('cdev' = sim.norm$cdev,
               'nGene' = sim.norm$nGene,
               'diffFraction' = sim.norm$diffFraction,
               'n_diffs_in_scale' = sim.norm$n_diffs_in_scale,
               'diffVarTotal' = sim.norm$diffVarTotal,
               'AUC' = auc,
               'TPR' = TP/length(which(sim$differential == 1)),
               'FDR' = FP / length(deg.idx),
               'FPR' = FP / length(which(sim$differential == 0))
               )
}) %>%
    do.call(rbind, .)

ggplot(performance.df) +
    facet_grid(. ~ nGene) +
    geom_point(aes(x=AUC,y=cdev, color=nGene, group=nGene), alpha = 0.3)
ggplot(performance.df) +
    facet_grid(. ~ nGene) +
    geom_point(aes(x=cdev,y=AUC, color=nGene, group=nGene), alpha = 0.3)
# ggplot(performance.df) +
#     geom_point(aes(x=diffVarTotal,y=cdev, color=(nGene), group=nGene), alpha = 0.3)
```

cdev vs reference set quality

```{r}
pl_cdev_vs_df.ngene = ggplot(performance.df) +
    geom_point(aes(x=diffFraction,y=cdev, color=nGene, group=nGene), alpha = 0.3) +
    theme_bw(base_size = THEME_BASE_SIZE)
print(pl_cdev_vs_df.ngene)
```
cdev vs DE performance

```{r}
pl_cdev_vs_auc.ngene = ggplot(performance.df) +
    geom_point(aes(x=cdev,y=AUC, color=nGene, group=nGene), alpha = 0.3) +
    theme_bw(base_size = THEME_BASE_SIZE)
print(pl_cdev_vs_auc.ngene)
```
## Varying fold-change

```{r}
NSAMPLES = 40
nGene = 20000
percentDE = 0.5 
foldChanges= c(1.5, 2, 3, 4, 5, 6, 7, 8) 
sims_f <- lapply(foldChanges, function(foldChange) {
    simulate.counts(nGene, NSAMPLES/2,NSAMPLES/2,fc = foldChange, percentDE = percentDE, completelyAsymm = TRUE)
}) %>%
    set_names(as.character(foldChanges))

sims_f.ref = lapply(foldChanges, function(foldChange) {
    sim = sims_f[[as.character(foldChange)]]
    normalize.by.refs(sim$counts, which(sim$differential == 0))
}) %>%
    set_names(as.character(foldChanges))
```

```{r}

sims_f.rand.norms =
    lapply(foldChanges, function(foldChange) {
        sim = sims_f[[as.character(foldChange)]]
        sim.norm.ref = sims_f.ref[[as.character(foldChange)]]
        trueRef.idx = which(sim$differential == 0)
        trueDif.idx = which(sim$differential == 1)
        NSCALE = min(length(trueDif.idx), length(trueRef.idx))
        mclapply(seq(1,NSCALE, floor(NSCALE/n_data_points)), mc.cores = 1, FUN = function(n_diff) {
            lapply(1:10, function(i) {
                rand.dif = sample(trueDif.idx, size = n_diff)
                rand.ref = sample(trueRef.idx, size = NSCALE - n_diff)
                rand.subset = union(rand.ref, rand.dif)
                raw = colSums(sim$counts[rand.subset,,drop= FALSE])
                sim.norm.i = normalize.by.refs(sim$counts, rand.subset)
                scalingFactors = raw / geom.mean(raw)
               
                mseRefs = mean((logFC(sim.norm.i[trueRef.idx,], sim$conditions$condition, log.base = 2, pseudocount = 1))^2) 
                return(list('scalingFactors' = scalingFactors,
                            'foldChange' = foldChange,
                            'rand.subset' = rand.subset,
                            'diffFraction' = n_diff/NSCALE,
                            'n_diffs_in_scale' = n_diff,
                            'diffVarTotal' = sum(var(sim$counts[rand.dif,])), # The total variance of the DE genes in the random set selected for scaling
                            'cdev' =  cdev(sim.norm.i, sim.norm.ref),
                            'mseRefs' = mseRefs))
            }) 
        }) %>%
        unlist(recursive = FALSE)
    }) %>%
    unlist(recursive = FALSE)
```

### DE tests

```{r, message=FALSE}
f_de.results = lapply(sims_f.rand.norms, function(sim.norm) {
    
    sim = sims_f[[as.character(sim.norm$foldChange)]]
    de.test.limma(sim$counts,
            group = sim$conditions$condition,
            norm.factors = sim.norm$scalingFactors) %>%
        limma::topTable(n = Inf, sort.by = 'none', adjust.method = 'BH')
})
```

```{r}
performance.df = lapply(1:length(sims_f.rand.norms), function(i) {
    
    sim.norm = sims_f.rand.norms[[i]]
    sim = sims_f[[as.character(sim.norm$foldChange)]]
    de = f_de.results[[i]]
    deg.idx = which(de$adj.P.Val < 0.01)
    
    auc = auroc(1-de$adj.P.Val, !!sim$differential)
    TP = length(intersect(which(sim$differential == 1), deg.idx))
    FP = length(setdiff(deg.idx, which(sim$differential == 1)))
    idx.diff = sim.norm$rand.subset[which(sim$differential[sim.norm$rand.subset] == 1)]
    
    data.frame('cdev' = sim.norm$cdev,
               'foldChange' = sim.norm$foldChange,
               'diffFraction' = sim.norm$diffFraction,
               'n_diffs_in_scale' = sim.norm$n_diffs_in_scale,
               'diffVarTotal' = sim.norm$diffVarTotal,
               'AUC' = auc,
               'TPR' = TP/length(which(sim$differential == 1)),
               'FDR' = FP / length(deg.idx),
               'FPR' = FP / length(which(sim$differential == 0))
               )
}) %>%
    do.call(rbind, .)
pl_cdev_vs_df.foldchange = ggplot(performance.df) +
    geom_point(aes(x=diffFraction,y=cdev, color=foldChange, group=foldChange), alpha = 0.5) +
    # guides(color=guide_legend(title = 'Fold-change')) +
    theme_bw(base_size = THEME_BASE_SIZE)
pl_df_vs_auc.foldchange = ggplot(performance.df) +
    geom_point(aes(x=diffFraction,y=AUC, color=foldChange, group=foldChange), alpha = 0.5) +
    # guides(color=guide_legend(title = 'Fold-change')) +
    theme_bw(base_size = THEME_BASE_SIZE)
pl_cdev_vs_auc.foldchange = ggplot(performance.df) +
    geom_point(aes(x=cdev,y=AUC, color=foldChange, group=foldChange), alpha = 0.5) +
    # guides(color=guide_legend(title = 'Fold-change')) +
    theme_bw(base_size = THEME_BASE_SIZE)

print(pl_cdev_vs_df.foldchange)
print(pl_cdev_vs_auc.foldchange)
print(pl_df_vs_auc.foldchange)
```

## Varying DE percentage

```{r}
NSAMPLES = 40
nGene = 10000 
percentDEs = c(1:9/10) 
foldChange = 4
sims_p <- lapply(percentDEs, function(percentDE) {
    simulate.counts(nGene, NSAMPLES/2,NSAMPLES/2,fc = foldChange, percentDE = percentDE, completelyAsymm = TRUE)
}) %>%
    set_names(as.character(percentDEs))

sims_p.ref = lapply(percentDEs, function(percentDE) {
    sim = sims_p[[as.character(percentDE)]]
    normalize.by.refs(sim$counts, which(sim$differential == 0))
}) %>%
    set_names(as.character(percentDEs))
```

Normalizing by a random set of equivalent size to the true reference set

```{r}
# diffFractions = c(0:50/50)

sims_p.rand.norms =
    lapply(percentDEs, function(percentDE) {
        sim = sims_p[[as.character(percentDE)]]
        sim.norm.ref = sims_p.ref[[as.character(percentDE)]]
        trueRef.idx = which(sim$differential == 0)
        trueDif.idx = which(sim$differential == 1)
        
        NSCALE = min(length(trueDif.idx), length(trueRef.idx))
        mclapply(seq(1,NSCALE, floor(NSCALE/n_data_points)), mc.cores = 1, FUN = function(n_diff) {
            lapply(1:10, function(i) {
                rand.dif = sample(trueDif.idx, size = n_diff)
                rand.ref = sample(trueRef.idx, size = NSCALE - n_diff)
                   
                rand.subset = union(rand.ref, rand.dif)
                raw = colSums(sim$counts[rand.subset,,drop= FALSE])
                sim.norm.i = normalize.by.refs(sim$counts, rand.subset)
                scalingFactors = raw / geom.mean(raw)
                
                return(list('scalingFactors' = scalingFactors,
                            'percentDE' = percentDE,
                            'rand.subset' = rand.subset,
                            'n_diffs_in_scale' = n_diff,
                            'diffFraction' = n_diff/NSCALE,
                            'diffVarTotal' = sum(var(sim$counts[rand.dif,])), # The total variance of the DE genes in the random set selected for scaling
                            'cdev' =  cdev(sim.norm.i, sim.norm.ref)))
            }) 
        }) %>%
        unlist(recursive = FALSE)
    }) %>%
    unlist(recursive = FALSE)
```

### DE tests

```{r, message=FALSE}
p_de.results = lapply(sims_p.rand.norms, function(sim.norm) {
    
    sim = sims_p[[as.character(sim.norm$percentDE)]]
    de.test.limma(sim$counts,
            group = sim$conditions$condition,
            norm.factors = sim.norm$scalingFactors) %>%
        limma::topTable(n = Inf, sort.by = 'none', adjust.method = 'BH')
})
```

```{r}
performance.df = lapply(1:length(sims_p.rand.norms), function(i) {
    
    sim.norm = sims_p.rand.norms[[i]]
    sim = sims_p[[as.character(sim.norm$percentDE)]]
    de = p_de.results[[i]]
    deg.idx = which(de$adj.P.Val < 0.01)
    
    auc = auroc(1-de$adj.P.Val, !!sim$differential)
    TP = length(intersect(which(sim$differential == 1), deg.idx))
    FP = length(setdiff(deg.idx, which(sim$differential == 1)))
    idx.diff = sim.norm$rand.subset[which(sim$differential[sim.norm$rand.subset] == 1)]
    
    data.frame('cdev' = sim.norm$cdev,
               'percentDE' = sim.norm$percentDE,
               'diffFraction' = sim.norm$diffFraction,
               'n_diffs_in_scale' = sim.norm$n_diffs_in_scale,
               'diffVarTotal' = sim.norm$diffVarTotal,
               'AUC' = auc,
               'TPR' = TP/length(which(sim$differential == 1)),
               'FDR' = FP / length(deg.idx),
               'FPR' = FP / length(which(sim$differential == 0))
               )
}) %>%
    do.call(rbind, .)
pl_cdev_vs_df.percentDE = ggplot(performance.df) +
    geom_point(aes(x=diffFraction,y=cdev, color=percentDE, group=percentDE), alpha = 0.3) +
    theme_bw(base_size = THEME_BASE_SIZE)

pl_cdev_vs_auc.percentDE = ggplot(performance.df) +
    geom_point(aes(x=cdev,y=AUC, color=percentDE, group=percentDE), alpha = 0.3) +
    theme_bw(base_size = THEME_BASE_SIZE)
print(pl_cdev_vs_df.percentDE)
print(pl_cdev_vs_auc.percentDE)
```

```{r,fig.width=5,fig.height=12}
ggplot(performance.df) +
    facet_grid(percentDE ~ .) +
    geom_point(aes(x=cdev,y=AUC, color=percentDE, group=percentDE), alpha = 0.3)
```


Normalizing by a random set of any size 

```{r}
# diffFractions = c(0:50/50)

sims_p.rand.norms =
    lapply(percentDEs, function(percentDE) {
        sim = sims_p[[as.character(percentDE)]]
        sim.norm.ref = sims_p.ref[[as.character(percentDE)]]
        trueRef.idx = which(sim$differential == 0)
        trueDif.idx = which(sim$differential == 1)
        
        NSCALE = min(length(trueDif.idx), length(trueRef.idx))
        mclapply(seq(1,NSCALE, floor(NSCALE/n_data_points)), mc.cores = 1, FUN = function(n_diff) {
            lapply(1:10, function(i) {
                rand.dif = sample(trueDif.idx, size = n_diff)
                rand.ref = sample(trueRef.idx, size = NSCALE - n_diff)
                   
                rand.subset = union(rand.ref, rand.dif)
                raw = colSums(sim$counts[rand.subset,,drop= FALSE])
                sim.norm.i = normalize.by.refs(sim$counts, rand.subset)
                scalingFactors = raw / geom.mean(raw)
                
                return(list('scalingFactors' = scalingFactors,
                            'percentDE' = percentDE,
                            'rand.subset' = rand.subset,
                            'n_diffs_in_scale' = n_diff,
                            'diffFraction' = n_diff/NSCALE,
                            'diffVarTotal' = sum(var(sim$counts[rand.dif,])), # The total variance of the DE genes in the random set selected for scaling
                            'cdev' =  cdev(sim.norm.i, sim.norm.ref)))
            }) 
        }) %>%
        unlist(recursive = FALSE)
    }) %>%
    unlist(recursive = FALSE)
```

```{r}
performance.df = lapply(1:length(sims_p.rand.norms), function(i) {
    
    sim.norm = sims_p.rand.norms[[i]]
    sim = sims_p[[as.character(sim.norm$percentDE)]]
    
    data.frame('cdev' = sim.norm$cdev,
               'percentDE' = sim.norm$percentDE,
               'diffFraction' = sim.norm$diffFraction,
               'n_diffs_in_scale' = sim.norm$n_diffs_in_scale,
               'diffVarTotal' = sim.norm$diffVarTotal
               )
}) %>%
    do.call(rbind, .)
```


```{r,fig.width=5,fig.height=12}
ggplot(performance.df) +
    facet_grid(percentDE ~ .) +
    geom_point(aes(x=n_diffs_in_scale,y=cdev, color=percentDE, group=percentDE), alpha = 0.3)
```


```{r}
ggplot(performance.df) +
    geom_point(aes(x=n_diffs_in_scale,y=cdev, color=percentDE, group=percentDE), alpha = 0.3)
ggplot(performance.df) +
    geom_point(aes(x=diffFraction,y=cdev, color=percentDE, group=percentDE), alpha = 0.3)
# ggplot(performance.df) +
#     geom_point(aes(x=diffVarTotal,y=cdev, color=(nGene), group=nGene), alpha = 0.3)
```

## Realistic simulations: heterogenous fold-changes

### Varying number of genes

```{r}
NSAMPLES = 40
nGenes = c(2000,10000, 20000, 50000)
percentDE = 0.5 
sims_rg <- lapply(nGenes, function(nGene) {
    simulate.counts.varyingfc(ngenes = nGene, nlibs1 = NSAMPLES %/% 2, nlibs2= NSAMPLES - (NSAMPLES %/% 2), percentDE = percentDE)
}) %>%
    set_names(as.character(nGenes))

sims_rg.ref = lapply(nGenes, function(nGene) {
    sim = sims_rg[[as.character(nGene)]]
    normalize.by.refs(sim$counts, which(sim$differential == 0))
}) %>%
    set_names(as.character(nGenes))
```
```{r}

sims_rg.rand.norms =
    lapply(nGenes, function(nGene) {
        sim = sims_rg[[as.character(nGene)]]
        sim.norm.ref = sims_rg.ref[[as.character(nGene)]]
        trueRef.idx = which(sim$differential == 0)
        trueDif.idx = which(sim$differential == 1)
        NSCALE = min(length(trueDif.idx), length(trueRef.idx))
        mclapply(seq(1,NSCALE, floor(NSCALE/n_data_points)), mc.cores = 1, FUN = function(n_diff) {
            lapply(1:10, function(i) {
                rand.dif = sample(trueDif.idx, size = n_diff)
                rand.ref = sample(trueRef.idx, size = NSCALE - n_diff)
                rand.subset = union(rand.ref, rand.dif)
                raw = colSums(sim$counts[rand.subset,,drop= FALSE])
                sim.norm.i = normalize.by.refs(sim$counts, rand.subset)
                scalingFactors = raw / geom.mean(raw)
                
                return(list('scalingFactors' = scalingFactors,
                            'nGene' = nGene,
                            'rand.subset' = rand.subset,
                            'diffFraction' = n_diff/NSCALE,
                            'n_diffs_in_scale' = n_diff,
                            'diffVarTotal' = sum(var(sim$counts[rand.dif,])), # The total variance of the DE genes in the random set selected for scaling
                            'cdev' =  cdev(sim.norm.i, sim.norm.ref)))
            }) 
        }) %>%
        unlist(recursive = FALSE)
    }) %>%
    unlist(recursive = FALSE)
```

#### DE tests

```{r, message=FALSE}
rg_de.results = lapply(sims_rg.rand.norms, function(sim.norm) {
    
    sim = sims_rg[[as.character(sim.norm$nGene)]]
    de.test.limma(sim$counts,
            group = sim$conditions$condition,
            norm.factors = sim.norm$scalingFactors) %>%
        limma::topTable(n = Inf, sort.by = 'none', adjust.method = 'BH')
})
```

```{r,fig.width=15,fig.height=4}
performance.df = lapply(1:length(sims_rg.rand.norms), function(i) {
    
    sim.norm = sims_rg.rand.norms[[i]]
    sim = sims_rg[[as.character(sim.norm$nGene)]]
    de = rg_de.results[[i]]
    deg.idx = which(de$adj.P.Val < 0.01)
    auc = auroc(1-de$adj.P.Val, !!sim$differential)
    TP = length(intersect(which(sim$differential == 1), deg.idx))
    FP = length(setdiff(deg.idx, which(sim$differential == 1)))
    idx.diff = sim.norm$rand.subset[which(sim$differential[sim.norm$rand.subset] == 1)]
    
    data.frame('cdev' = sim.norm$cdev,
               'nGene' = sim.norm$nGene,
               'diffFraction' = sim.norm$diffFraction,
               'n_diffs_in_scale' = sim.norm$n_diffs_in_scale,
               'diffVarTotal' = sim.norm$diffVarTotal,
               'AUC' = auc,
               'TPR' = TP/length(which(sim$differential == 1)),
               'FDR' = FP / length(deg.idx),
               'FPR' = FP / length(which(sim$differential == 0))
               )
}) %>%
    do.call(rbind, .)

ggplot(performance.df) +
    facet_grid(. ~ nGene) +
    geom_point(aes(x=AUC,y=cdev, color=nGene, group=nGene), alpha = 0.3) +
    theme_bw(base_size = THEME_BASE_SIZE)
ggplot(performance.df) +
    facet_grid(. ~ nGene) +
    geom_point(aes(x=cdev,y=AUC, color=nGene, group=nGene), alpha = 0.3) +
    theme_bw(base_size = THEME_BASE_SIZE)
```

```{r}
p2_cdev_vs_df.ngene = ggplot(performance.df) +
    geom_point(aes(x=diffFraction,y=cdev, color=nGene, group=nGene), alpha = 0.3) +
    theme_bw(base_size = THEME_BASE_SIZE)
print(p2_cdev_vs_df.ngene)
```

```{r}
p2_cdev_vs_auc.ngene = ggplot(performance.df) +
    geom_point(aes(x=cdev,y=AUC, color=nGene, group=nGene), alpha = 0.3) +
    theme_bw(base_size = THEME_BASE_SIZE)
print(p2_cdev_vs_auc.ngene)
```

### Varying DE percentage

```{r}
NSAMPLES = 40
nGene = 10000 
percentDEs = c(1:9/10) 
sims_rp <- lapply(percentDEs, function(percentDE) {
    simulate.counts.varyingfc(ngenes = nGene, nlibs1 = NSAMPLES %/% 2, nlibs2= NSAMPLES - (NSAMPLES %/% 2), percentDE = percentDE)
}) %>%
    set_names(as.character(percentDEs))

sims_rp.ref = lapply(percentDEs, function(percentDE) {
    sim = sims_rp[[as.character(percentDE)]]
    normalize.by.refs(sim$counts, which(sim$differential == 0))
}) %>%
    set_names(as.character(percentDEs))
```

Normalizing by a random set of equivalent size to the true reference set

```{r}
# diffFractions = c(0:50/50)

sims_rp.rand.norms =
    lapply(percentDEs, function(percentDE) {
        sim = sims_rp[[as.character(percentDE)]]
        sim.norm.ref = sims_rp.ref[[as.character(percentDE)]]
        trueRef.idx = which(sim$differential == 0)
        trueDif.idx = which(sim$differential == 1)
        
        NSCALE = min(length(trueDif.idx), length(trueRef.idx))
        mclapply(seq(1,NSCALE, floor(NSCALE/n_data_points)), mc.cores = 1, FUN = function(n_diff) {
            lapply(1:10, function(i) {
                rand.dif = sample(trueDif.idx, size = n_diff)
                rand.ref = sample(trueRef.idx, size = NSCALE - n_diff)
                   
                rand.subset = union(rand.ref, rand.dif)
                raw = colSums(sim$counts[rand.subset,,drop= FALSE])
                sim.norm.i = normalize.by.refs(sim$counts, rand.subset)
                scalingFactors = raw / geom.mean(raw)
                
                return(list('scalingFactors' = scalingFactors,
                            'percentDE' = percentDE,
                            'rand.subset' = rand.subset,
                            'n_diffs_in_scale' = n_diff,
                            'diffFraction' = n_diff/NSCALE,
                            'diffVarTotal' = sum(var(sim$counts[rand.dif,])), # The total variance of the DE genes in the random set selected for scaling
                            'cdev' =  cdev(sim.norm.i, sim.norm.ref)))
            }) 
        }) %>%
        unlist(recursive = FALSE)
    }) %>%
    unlist(recursive = FALSE)
```

#### DE tests

```{r, message=FALSE}
rp_de.results = lapply(sims_rp.rand.norms, function(sim.norm) {
    
    sim = sims_rp[[as.character(sim.norm$percentDE)]]
    de.test.limma(sim$counts,
            group = sim$conditions$condition,
            norm.factors = sim.norm$scalingFactors) %>%
        limma::topTable(n = Inf, sort.by = 'none', adjust.method = 'BH')
})
```

```{r}
performance.df = lapply(1:length(sims_rp.rand.norms), function(i) {
    
    sim.norm = sims_rp.rand.norms[[i]]
    sim = sims_rp[[as.character(sim.norm$percentDE)]]
    de = rp_de.results[[i]]
    deg.idx = which(de$adj.P.Val < 0.01)
    
    auc = auroc(1-de$adj.P.Val, !!sim$differential)
    TP = length(intersect(which(sim$differential == 1), deg.idx))
    FP = length(setdiff(deg.idx, which(sim$differential == 1)))
    idx.diff = sim.norm$rand.subset[which(sim$differential[sim.norm$rand.subset] == 1)]
    
    data.frame('cdev' = sim.norm$cdev,
               'percentDE' = sim.norm$percentDE,
               'diffFraction' = sim.norm$diffFraction,
               'n_diffs_in_scale' = sim.norm$n_diffs_in_scale,
               'diffVarTotal' = sim.norm$diffVarTotal,
               'AUC' = auc,
               'TPR' = TP/length(which(sim$differential == 1)),
               'FDR' = FP / length(deg.idx),
               'FPR' = FP / length(which(sim$differential == 0))
               )
}) %>%
    do.call(rbind, .)
p2_cdev_vs_df.percentDE = ggplot(performance.df) +
    geom_point(aes(x=diffFraction,y=cdev, color=percentDE, group=percentDE), alpha = 0.3) +
    theme_bw(base_size = THEME_BASE_SIZE)
# ggplot(performance.df) +
#     geom_point(aes(x=diffFraction,y=AUC, color=percentDE, group=percentDE), alpha = 0.3) +
#     theme_bw(base_size = THEME_BASE_SIZE)

p2_cdev_vs_auc.percentDE = ggplot(performance.df) +
    geom_point(aes(x=cdev,y=AUC, color=percentDE, group=percentDE), alpha = 0.3) +
    theme_bw(base_size = THEME_BASE_SIZE)
print(p2_cdev_vs_df.percentDE)
print(p2_cdev_vs_auc.percentDE)
ggplot(performance.df) +
    facet_wrap('percentDE') +
    geom_point(aes(x=cdev,y=AUC, color=percentDE, group=percentDE), alpha = 0.3) +
    theme_bw(base_size = THEME_BASE_SIZE)
```

```{r}
ggplot(performance.df) +
    geom_point(aes(x=diffFraction,y=cdev, color=percentDE, group=percentDE), alpha = 0.3) +
    theme_bw(base_size = THEME_BASE_SIZE)
```
```{r}
ggplot(performance.df) +
    geom_point(aes(x=diffVarTotal,y=cdev, color=percentDE, group=percentDE), alpha = 0.3)
```

## Combining plots for publication figure


```{r,fig.width=8,fig.height=15}
egg::ggarrange(pl_cdev_vs_df.ngene + theme(legend.position = 'none',axis.title.x=element_blank()),pl_cdev_vs_auc.ngene + theme(axis.title.x=element_blank()),
               pl_cdev_vs_df.percentDE + theme(legend.position = 'none',axis.title.x=element_blank()),pl_cdev_vs_auc.percentDE  + theme(axis.title.x=element_blank()),
               pl_cdev_vs_df.foldchange + theme(legend.position = 'none',axis.title.x=element_blank()),pl_cdev_vs_auc.foldchange + theme(axis.title.x=element_blank()),
               p2_cdev_vs_df.ngene + theme(legend.position = 'none',axis.title.x=element_blank()),p2_cdev_vs_auc.ngene + theme(axis.title.x=element_blank()),
               p2_cdev_vs_df.percentDE + theme(legend.position = 'none'),p2_cdev_vs_auc.percentDE,
               ncol = 2)
```

